AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology

### 2013-06-03 front-matter.tex
retirados hifens de auto-regulationS, mudado meu email, arrumado unfold de IFSC, colocados pts finais nos rodapes

\author[1,4]{Renato Fabbri\thanks{renato.fabbri@gmail.com}}
\author[2,4]{Ricardo Fabbri\thanks{rfabbri@iprj.uerj.br}}
\author[1,4]{Vilson Vieira\thanks{vilson@void.cc}}
\author[4]{Alexandre Negr\~{a}o}
\author[4]{Lucas Zambianchi}
\author[4]{Marcos~Mendon\c{c}a}
\author[1,3]{Daniel Penalva}
\author[4]{Danilo Shiga}
\affil[1]{\small Physics Insitute at S\~{a}o Carlos (IFSC), University of S\~{a}o Paulo (USP), S\~ao Carlos, Brazil}
\affil[2]{\small Polytechnic Institute (IPRJ), State University of Rio de Janeiro (UERJ), Nova Friburgo, Brazil}
\affil[3]{\small  Institute for Theoretical Physics (IFT), Universidade Estadual Paulista (UNESP), S\~ao Paulo, Brazil}
\affil[4]{\small \url{LabMacambira.sourceforge.net}}

@@

\begin{abstract}
We present a new self-regulating methodology for coordinating distributed team
work called Algorithmic Auto-regulation (AA), based on recent social networking
concepts and individual merit. Team members take on an egalitarian role, and
stay voluntarily logged into so-called AA sessions for part of their time (e.g.\
2 hours per day), during which they create periodical logs --- short text
sentences --- they wish to share about their activity with the team. These logs
are publicly aggregated in a Website and are peer-validated after the end of a
session, as in code review.  A short screencast is ideally recorded at the end
of each session to make AA logs more understandable.  This methodology has shown
to be well-suited for increasing the efficiency of distributed teams working on
what is called Global Software Development (GSD), as observed in our experience
in actual real-world situations.  This efficiency boost is mainly achieved through 1)
built-in asyncrhonous on-demand communication, documentation of work products
and processes, and 2) reduced need for central management, meetings or
time-consuming reports. Hence, the AA methodology legitimizes and facilitates
the activities of a distributed software team.  It thus enables other entities
to have a solid means to fund these activities, allowing for new and concrete
business models to emerge for very distributed software development. AA has been
proposed, at its core, as a way of sustaining self-replicating hacker
initiatives. These claims are discussed in a real case-study of running a
distributed free software hacker team called Lab Macambira.  
\end{abstract}

>>>>

\author[1,4]{Renato Fabbri\thanks{fabbri@usp.br}}
\author[2,4]{Ricardo Fabbri\thanks{rfabbri@iprj.uerj.br}}
\author[1,4]{Vilson Vieira\thanks{vilson@void.cc}}
\author[4]{Alexandre Negr\~{a}o}
\author[4]{Lucas Zambianchi}
\author[4]{Marcos~Mendon\c{c}a}
\author[1,3]{Daniel Penalva}
\author[4]{Danilo Shiga}
\affil[1]{\small S\~ao Carlos Institute of Physics (IFSC), University of S\~{a}o Paulo (USP), S\~ao Carlos, Brazil}
\affil[2]{\small Polytechnic Institute (IPRJ), State University of Rio de Janeiro (UERJ), Nova Friburgo, Brazil}
\affil[3]{\small  Institute for Theoretical Physics (IFT), Universidade Estadual Paulista (UNESP), S\~ao Paulo, Brazil}
\affil[4]{\small \url{LabMacambira.sourceforge.net}}

@@

\begin{abstract}
We present a new self-regulating methodology for coordinating distributed team
work called Algorithmic Autoregulation (AA), based on recent social networking
concepts and individual merit. Team members take on an egalitarian role, and
stay voluntarily logged into so-called AA sessions for part of their time (e.g.\
2 hours per day), during which they create periodical logs --- short text
sentences --- they wish to share about their activity with the team. These logs
are publicly aggregated in a Website and are peer-validated after the end of a
session, as in code review.  A short screencast is ideally recorded at the end
of each session to make AA logs more understandable.  This methodology has shown
to be well-suited for increasing the efficiency of distributed teams working on
what is called Global Software Development (GSD), as observed in our experience
in actual real-world situations.  This efficiency boost is mainly achieved through 1)
built-in asyncrhonous on-demand communication, documentation of work products
and processes, and 2) reduced need for central management, meetings or
time-consuming reports. Hence, the AA methodology legitimizes and facilitates
the activities of a distributed software team.  It thus enables other entities
to have a solid means to fund these activities, allowing for new and concrete
business models to emerge for very distributed software development. AA has been
proposed, at its core, as a way of sustaining self-replicating hacker
initiatives. These claims are discussed in a real case-study of running a
distributed free software hacker team called Lab Macambira.  
\end{abstract}

### 2013-06-03 paper.tex
todo discussion about epistemological shift of academy with the AA methodology productive to processual
retirados hifens de auto-regulationS, mudado meu email, arrumado unfold de IFSC, colocados pts finais nos rodapes

One of the defining features of modern times is the widening geographical
distribution of software teams~\cite{last2003} creating what is called Global
Software Development (GSD)~\cite{german2003,Fryer,Begel}.  
An example is the free software movement. Projects and institutions like Mozilla
Foundation has several employees and thousands of voluntary developers
distributed across many countries. The same is true for GNOME~\cite{german2003},
OpenBSD, MySQL or Apache Software Foundation, to cite just a 
few of the most active projects.\footnote{Ohloh, the open source network, has a more complete
and constantly updated list of the most active projects on-line at
\url{www.ohloh.net}} Beyond the free and open software community, GSD has a growing
popularity in every niche of the software industry as a whole, even among those
distributing their software with proprietary licenses. This phenomenon is
attributed to a variety of factors such as a larger labor pool, natural
globalization of software companies and foundations or even the premise of
cheaper cost of production~\cite{komi2005}.

@@

Easy and effective GSD team management is the main purpose of the AA
methodology. The proposed methodology was applyed to a group of 9 developers in
July of 2011, 3 of which are coauthors of the present work, on what was named Lab
Macambira.\footnote{LabMacambira.sf.net: \url{http://labmacambira.sf.net}} The
main objective of the team was to work on an array of strategic free software
projects in the broad audiovisual and web categories, contributing directly to
their development, submitting bug patches or committing new features to their
source code.

@@

In one month, each developer officially contributed to one or many free software
projects. Many developers started the initiation training with no knowledge of what
is free software and ended that period becoming a free software developer.
During that month, the same team of trainees also developed the first version of the AA
system and used AA to manage their activities, even while developing
the other aforementioned free software projects. Thus, AA and the associated
software system was tested, prototyped, and developed in close contact with
actual practice.  The source code of AA --- both the client that sends the
logs and the AA Web server --- is public available as free software\footnote{AA source
  code:
  \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/aa}}
and all the actual AA log data of the entire team of Lab Macambira from 2011 to
the present time is also on-line\footnote{Logs of AA sessions:
  \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/paainel}}.

After the initial training period of 1 month, the initiated `Macambiras'
worked during 6 additional months on a large range of free software projects,
divided into work groups --- each work group focusing on a specific theme like
video, audio and web --- and funded by contracts, freelance, and support of the
Pont\~{a}o N\'{o}s Digitais\footnote{\url{http://nosdigitais.teia.org.br}}.
Table~\ref{tabela:criados} has a list of \emph{new} free software applications
created by `Lab Macambira' since July 2011.

@@

\begin{table}
    \caption{Software projects created by Lab Macambira since July
      2011 with a short description and the technologies involved (\eg,
      programming languages or frameworks). It is
      interesting to note the heterogeneity of projects and their areas
      of application.}
    \footnotesize{\begin{tabular}{|l|p{5cm}|l|}
        \hline
        Application & Description & Technologies \\ 
        \hline \hline
        AA            & Algorithmic Auto-regulation      & Python, PHP \\
        \hline
        \'{A}gora Communs & System for on-line deliberations & PHP \\
        \hline
        SIP           & Scilab Image Processing toolbox & C, Scilab \\
        \hline
        Animal        & An Imaging Library              & C \\
        \hline
        TeDi          & Test Framework for Distance Transform
        Algorithms & C, Shellscript, Scilab \\
        \hline
        Macambot      & Multi-use IRC Bot               & Python \\
        \hline
        Confer\^{e}ncia Permanente & Platform for the permanent
        conference of the rights of minors & PHP, JavaScript \\
        \hline
        CPC           & Center for accounting of the Brazilian culture
        representation groups & Python, Django \\
        \hline
        Timeline      & Interactive time lines on the Web & JavaScript
        \\
        \hline
        Imagemap      & Interactive marking for on-line photos &
        JavaScript \\
        \hline
        ABT           & Program for real-time execution and musical
        rhythmic analysis & Python \\
        \hline
        EKP           & Emotional Kernel Panic & Python, ChucK \\
        \hline
        SOS           & Aggregation and diffusion of popular and native
        knowledge about health & Python, Django \\
        \hline
        Creative Economy & Platform for creative, collaborative and
        solidarity economy of the culture hubs and cultural entities &
        Python, Django \\
        \hline
        OpenID Integration & Adaptations to existing software for
        unified login through OpenID & PHP \\
        \hline
        pAAinel & Dashboard for the real-time visualization of Lab
        Macambira activity & Python, Django \\
        \hline
        Georef & Collection of scripts to be used as reference, which
        aims to be a GIS platform to map public data of use to
        citizens & Python, Django \\
        \hline
        AirHackTable & Software for an instrument which generates
        sound from flying origami tracked by webcams & Puredata,
        C++, Scilab \\
        \hline
        \end{tabular}}
    \label{tabela:criados}
\end{table}

@@

Authors have analyzed data and gained practical experience on
the use of AA to auto-regulate the work of Lab Macambira, a group of free
software developers from Brazil. Since July of 2011 the group has contributed
and created new free and open source software for a vast number of additional applications.

The AA methodology is not restricted to software development, even though it was
designed for the latter. As of this writing there is an entertainment studio,
Pula Pirata\footnote{ \url{www.pulapirata.com}}, that has been using AA to
manage their creative activities.  Other people with no software background,
like social scientists, musicians and activists also have been using AA and
contributing for its broader improvement.

@@

>>>>

One of the defining features of modern times is the widening geographical
distribution of software teams~\cite{last2003} creating what is called Global
Software Development (GSD)~\cite{german2003,Fryer,Begel}.  
An example is the free software movement. Projects and institutions like Mozilla
Foundation has several employees and thousands of voluntary developers
distributed across many countries. The same is true for GNOME~\cite{german2003},
OpenBSD, MySQL or Apache Software Foundation, to cite just a 
few of the most active projects.\footnote{Ohloh, the open source network, has a more complete
and constantly updated list of the most active projects on-line at
\url{www.ohloh.net}.} Beyond the free and open software community, GSD has a growing
popularity in every niche of the software industry as a whole, even among those
distributing their software with proprietary licenses. This phenomenon is
attributed to a variety of factors such as a larger labor pool, natural
globalization of software companies and foundations or even the premise of
cheaper cost of production~\cite{komi2005}.

@@

Easy and effective GSD team management is the main purpose of the AA
methodology. The proposed methodology was applyed to a group of 9 developers in
July of 2011, 3 of which are coauthors of the present work, on what was named Lab
Macambira.\footnote{LabMacambira.sf.net: \url{http://labmacambira.sf.net}.} The
main objective of the team was to work on an array of strategic free software
projects in the broad audiovisual and web categories, contributing directly to
their development, submitting bug patches or committing new features to their
source code.

@@

In one month, each developer officially contributed to one or many free software
projects. Many developers started the initiation training with no knowledge of what
is free software and ended that period becoming a free software developer.
During that month, the same team of trainees also developed the first version of the AA
system and used AA to manage their activities, even while developing
the other aforementioned free software projects. Thus, AA and the associated
software system was tested, prototyped, and developed in close contact with
actual practice.  The source code of AA --- both the client that sends the
logs and the AA Web server --- is public available as free software\footnote{AA source
  code:
  \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/aa}.}
and all the actual AA log data of the entire team of Lab Macambira from 2011 to
the present time is also on-line\footnote{Logs of AA sessions:
  \url{http://labmacambira.git.sourceforge.net/git/gitweb.cgi?p=labmacambira/paainel}.}.

After the initial training period of 1 month, the initiated `Macambiras'
worked during 6 additional months on a large range of free software projects,
divided into work groups --- each work group focusing on a specific theme like
video, audio and web --- and funded by contracts, freelance, and support of the
Pont\~{a}o N\'{o}s Digitais\footnote{\url{http://nosdigitais.teia.org.br}.}.
Table~\ref{tabela:criados} has a list of \emph{new} free software applications
created by `Lab Macambira' since July 2011.

@@

\begin{table}
    \caption{Software projects created by Lab Macambira since July
      2011 with a short description and the technologies involved (\eg,
      programming languages or frameworks). It is
      interesting to note the heterogeneity of projects and their areas
      of application.}
    \footnotesize{\begin{tabular}{|l|p{5cm}|l|}
        \hline
        Application & Description & Technologies \\ 
        \hline \hline
        AA            & Algorithmic Autoregulation      & Python, PHP \\
        \hline
        \'{A}gora Communs & System for on-line deliberations & PHP \\
        \hline
        SIP           & Scilab Image Processing toolbox & C, Scilab \\
        \hline
        Animal        & An Imaging Library              & C \\
        \hline
        TeDi          & Test Framework for Distance Transform
        Algorithms & C, Shellscript, Scilab \\
        \hline
        Macambot      & Multi-use IRC Bot               & Python \\
        \hline
        Confer\^{e}ncia Permanente & Platform for the permanent
        conference of the rights of minors & PHP, JavaScript \\
        \hline
        CPC           & Center for accounting of the Brazilian culture
        representation groups & Python, Django \\
        \hline
        Timeline      & Interactive time lines on the Web & JavaScript
        \\
        \hline
        Imagemap      & Interactive marking for on-line photos &
        JavaScript \\
        \hline
        ABT           & Program for real-time execution and musical
        rhythmic analysis & Python \\
        \hline
        EKP           & Emotional Kernel Panic & Python, ChucK \\
        \hline
        SOS           & Aggregation and diffusion of popular and native
        knowledge about health & Python, Django \\
        \hline
        Creative Economy & Platform for creative, collaborative and
        solidarity economy of the culture hubs and cultural entities &
        Python, Django \\
        \hline
        OpenID Integration & Adaptations to existing software for
        unified login through OpenID & PHP \\
        \hline
        pAAinel & Dashboard for the real-time visualization of Lab
        Macambira activity & Python, Django \\
        \hline
        Georef & Collection of scripts to be used as reference, which
        aims to be a GIS platform to map public data of use to
        citizens & Python, Django \\
        \hline
        AirHackTable & Software for an instrument which generates
        sound from flying origami tracked by webcams & Puredata,
        C++, Scilab \\
        \hline
        \end{tabular}}
    \label{tabela:criados}
\end{table}

@@

Authors have analyzed data and gained practical experience on
the use of AA to autoregulate the work of Lab Macambira, a group of free
software developers from Brazil. Since July of 2011 the group has contributed
and created new free and open source software for a vast number of additional applications.

The AA methodology is not restricted to software development, even though it was
designed for the latter. As of this writing there is an entertainment studio,
Pula Pirata\footnote{ \url{www.pulapirata.com}.}, that has been using AA to
manage their creative activities.  Other people with no software background,
like social scientists, musicians and activists also have been using AA and
contributing for its broader improvement.

@@

%\todo{ Discussao : AA como mudanca de epistemologia academica, vinculacao do valor processual atraves do AA, citacao nos shouts (durante trabalho academico), validacao de citacoes atraves de validacao do log de AA + revisao do processo academico presento no log em algum nivel de compreensao midiatica}
